---
format: 
  revealjs:
    slide-number: false
    progress: false
    code-overflow: wrap
    chalkboard: true
---

## Talking points

- Why random sampling?

- Estimation vs. inference

- Plug-in principle

- Estimation theory vs. *Asymptotic* estimation theory

- Confidence intervals vs. hypotheses testing

## Ingredients for statistical inference

::: incremental
1. Estimand $\theta$

2. Estimator $\widehat \theta$
:::

## Ingredients for statistical inference

1. Estimand $\mu$

2. Estimator $\widehat \mu$

## Ingredients for statistical inference

1. Estimand $\mu$

2. Estimator $\widehat \mu$

3. Data $X$

## Ingredients for statistical inference

1. Estimand $\mu$

2. Estimator $\widehat \mu$

3. Data $X$

4. Estimate $\overline X$

## Ingredients for statistical inference

1. Estimand $\mu$

2. Estimator $\widehat \mu$

3. Data $X$

4. Estimate $\overline X$

5. Measure of uncertainty (confidence intervals, hypothesis testing)


. . .

::: {style="text-align: center"}

$X \rightarrow \overline{X} \rightarrow \hat{\mu} \xrightarrow{\text{hopefully!}} \mu$

:::

## Data

```{r setup, include=FALSE}
# Global options for the rendering behavior of all subsequent code chunks
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE,
                      tidy = TRUE,
                      tidy.opts=list(width.cutoff=80))
```

```{r, eval = FALSE}
remotes::install_github("kjhealy/gssr")

library(gssr)

gss22 = gss_get_yr(2022)

# Question 202a. Tell me if you agree or disagree with this statement: Most men are better suited emotionally for politics than are most women.
# Find a more granular question!
x = gss22$fepol

table(x)
```

```{r, echo = FALSE, cache = TRUE}
library(gssr)
gss22 = gss_get_yr(2022)
x = gss22$fepol
table(x)
```








