---
title: "Projects and Tables"
format: 
  revealjs:
    slide-number: false
    progress: false
    chalkboard: true
    code-overflow: wrap
    
---

```{r setup, include=FALSE}
# Global options for the rendering behavior of all subsequent code chunks
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE,
                      fig.pos = "center")
```

## Presenting Data using R

```{r}
#| include: false
library(tidyverse)
```

Let's, say, work with our favorite `mtcars` data. 

<br>

What do you think of this presentation?

```{r}
head(mtcars)
```

## Improving the presentation

`KableExtra` can help you with this. Let's try it out.

```{r}
library(kableExtra)

mtcars_kable = mtcars %>%
  head(n = 3) %>%
  kbl()

mtcars_kable
```

## Custumizing the output

::: {.panel-tabset}
## Paper

```{r}
mtcars_kable %>%
  kable_paper()
```


## Classic

```{r}
mtcars_kable %>%
  kable_classic()
```

## Minimal

```{r}
mtcars_kable %>%
  kable_minimal()
```

## Material

```{r}
mtcars_kable %>%
  kable_material()
```

:::


## Presenting data we want

Let's say, we want to present average and median of `mpg` variable.

```{r}
descriptive_statistics = mtcars %>%
  summarize(mpg_average = mean(mpg),
            mpg_median = median(mpg))
```

<br>

```{r}
descriptive_statistics
```

## Making the data accessible

::: {.panel-tabset}
## Wide format

```{r}
descriptive_statistics %>%
  rename(`Average Miles Per Gallon` = mpg_average,
         `Median Miles Per Gallon` = mpg_median) %>%
  kbl() %>%
  kable_styling()
```


## Long format

```{r}
descriptive_statistics %>%
  pivot_longer(cols = c(mpg_average, mpg_median), 
               names_to = "Statistic",
               values_to = "Value") %>%
  mutate(Statistic = case_when(Statistic == "mpg_average" ~ "Average Miles Per Gallon",
                               Statistic == "mpg_median" ~ "Median Miles Per Gallon")) %>%
  kbl() %>%
  kable_styling()
```

:::


## Projects in R and paths

Absolute paths

```{r, eval=FALSE}
"G:/PS403 Stats and Probability/Labs/Lab0/data/data.csv"
```

<br>

Relative paths (with working directory set)

```{r, eval=FALSE}
"data/data.csv"
```

## Projects

Projects set your working directory to specific folder on your laptop

[How to find a directory on MacOS](https://support.apple.com/guide/mac-help/get-file-folder-and-disk-information-on-mac-mchlp1774)

<br>

Let's practice with happiness report 2024!
