---
title: "Lab 2: Quantities of interest"
author: 
  - FirstName LastName
  - <YourEmail@u.northwestern.edu>
date: October 14, 2024
date-format: "**[D]ue:** MMMM D, YYYY"
format: 
     pdf:
       documentclass: article
fontsize: 12pt
urlcolor: blue
number-sections: true
geometry:
      - left=1in
      - right=1in
      - top=1in
      - bottom=1in
header-includes:
    - \usepackage{setspace}
    - \doublespacing
    - \usepackage{float}
    - \floatplacement{figure}{t}
    - \floatplacement{table}{t}
    - \usepackage{flafter}
    - \usepackage[T1]{fontenc}
    - \usepackage[utf8]{inputenc}
    - \usepackage{ragged2e}
    - \usepackage{booktabs}
    - \usepackage{amsmath}
    - \usepackage{url}
    - \usepackage{fvextra}
    - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
# Global options for the rendering behavior of all subsequent code chunks
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE,
                      tidy = TRUE,
                      tidy.opts=list(width.cutoff=80))
```

# Setup {.unnumbered}

Before starting, include all the relevant packages in the code chunk below.

```{r}
library(tidyverse)
```

# Overview {.unnumbered}

AM chapter 2 introduces many summary features of random variables. We have also discussed how one nice feature of random variables is that we can treat them as a functions. This means the quantities introduced in the chapter also have an analogue that we can compute with actual data. Knowing this will be useful later to visualize the properties of different *estimators*.

# Create data

The following code simulates a data set.

```{r}
# Remember to use set.seed() so that results remain the same
data = data.frame(
  x = rnorm(n = 10000, mean = 5, sd = 2),
  y = rbinom(10000, size = 1, prob = 0.6)
)
```

## What does the object `data` contain?

## How is `x` generated? What does each argument mean?

## How is `y` generated? What does each argument mean?


# Summary quantities of interest

The code chunks below represent many quantities of interest we can compute. Under each chunk, Explain:

- What are we computing?
- Does this quantity have a name?
- Why is this quantity useful/important?
- Is there an easier way to calculate this in base R? What about a package? If so, show it

##

```{r, eval = FALSE}
sum(data$x)/length(data$x)
```

##

```{r, eval = FALSE}
sum(data$x^2)/length(data$x)
```

##

```{r, eval = FALSE}
sum((data$x - 3)^2)/length(data$x)
```

##

```{r, eval = FALSE}
sum((data$x - 0)^2)/length(data$x)
```

##

```{r, eval = FALSE}
with(data,
     sum((x - (sum(x)/length(x)))^2)/length(x)
     )
```


##

```{r, eval = FALSE}
with(data,
     sum(x^2/length(x)) -
       (sum(x)/length(x))^2
     )
```

##

```{r, eval = FALSE}
with(data, mean(x^2) - mean(x)^2)
```

##

```{r, eval = FALSE}
with(data,
     sqrt(mean(x^2) - mean(x)^2))
```

##

```{r, eval = FALSE}
with(data,
     sqrt(mean((x+5)^2) - mean(x+5)^2))
```

##

```{r, eval = FALSE}
with(data,
     sum((x - (sum(x)/length(x)))^3)/length(x)
     )
```


##

```{r, eval = FALSE}
with(data,
     sum((x - (sum(x)/length(x)))^4)/length(x)
     )
```


##

```{r, eval = FALSE}
with(data,
     (sum((x - (sum(x)/length(x)))^3)/length(x)) /
       (sqrt(mean(x^2) - mean(x)^2)^3)
     )
```

##

```{r, eval = FALSE}
with(data,
     (sum((x - (sum(x)/length(x)))^4)/length(x)) /
       (sqrt(mean(x^2) - mean(x)^2)^4)
     )
```


##

```{r, eval = FALSE}
with(data,
     (mean(x^2) - mean(x)^2) +
       (mean(x) - 3)^2
     )
```

##

```{r, eval = FALSE}
with(data, cov(x,y))

with(data,
     mean((x - mean(x)) * (y - mean(y)))
     )

```

##

```{r}
with(data,
     mean(x*y) - mean(x)*mean(y)
     )
```

##

```{r, eval = FALSE}
# Correlation
```

# Do again with any data set you like



